Expectations bias moral evaluations

### 2018-06-29 paper/paper.tex
added norming data

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Expectations bias moral evaluations},
            pdfauthor={Derek Powell and Zachary Horne},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Expectations bias moral evaluations}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Derek Powell and Zachary Horne}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
    \date{}
    \predate{}\postdate{}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

On the evening of November 13th, 2015, a terrorist attack in Paris left
130 people dead and injured over 300 more. In the aftermath, millions
took to Twitter to express their shock, horror, and outrage at this
tragedy under hashtags like \texttt{\#parisattacks} and
\texttt{\#jesuisparis}. Yet, most of those mourning had little to say 15
hours earlier, when another tragic attack killed at least 43 people in
Beirut. Several factors are surely at play in these different reactions
(construal-level theory, {\textbf{???}}; e.g.~group affiliations,
Brewer, 1999), yet one potentially fundamental factor has gone
unmentioned: the fact that the Paris attack was more surprising than the
attack in Beirut. In contrast to France, Lebanon had experienced dozens
of terrorist bombings and attacks in recent years. Consequently, Beirut
may seem to many like the sort of place where ``these things happen,''
whereas Paris is perceived as being stable and safe.

We can see in everyday experience that people's evaluations of events
often depend on their expectations about those events. There is often
disappointment when events fail to meet expectations, and there is a
special thrill to having one's expectations exceeded. Anecdotally, these
forces seem to drive people's tendencies to root for the underdog, hold
surprise parties, and foreshadow bad news to ease its delivery (Bell,
1985). Indeed, laboratory studies suggest that expectations play an
important role in people's evaluations of the utility of an event. For
instance, Mellers and colleagues (1997) found that expectations
influenced affective reactions during a gambling task: Given a gamble
with a 10\% chance to win \$30 and 90\% chance to win \$0, participants
felt little disappointment with the \$0 outcome, but were considerably
more excited when they won \$30. Conversely, given a gamble with a 90\%
chance of winning \$60 and 10\% chance of winning \$0, participants were
disappointed with the \$0 outcome and showed more muted enjoyment of the
\$60 outcome. In fact, in gambles similar to these, Mellers and
colleagues (1997) found that people were happier with the smaller
unexpected gain than with the larger but more expected gain (also see
Shepperd \& Mcnulty, 2002).

Just as expectations affect the utility of simple gambles, we suspect
they are equally likely to shape how people react to morally harmful
events, such as acts of terrorism (Waldmann Cite and Rai and Fiske cite
--domain general peeps; us also). However, unlike in the context of
gambles, in these contexts the effects of expectations on evaluations
may have harmful consequences. When events are shocking, we predict that
people will perceive them as more severe and consequently be roused to
action. In contrast, when events harm victims who are generally
considered to be at greater risk--the poor, sick, or those living in
unstable regions of the world, we predicted that reactions may be more
muted. However, prior research has yet to articulate whether these same
effects emerge in the moral domain and the implications of this
possibility. For instance, it is unclear if these same mechanisms can
help account for the divergent reactions people have to terrorism in
France and Beirut. Further, people's moral evaluations are in fact
impacted by expectations what is their source and what does that tell us
about how to overcome the biases they may impose?

A number of researchers have sought to develop theories of
disappointment--the psychological reactions that result when experiences
fail to meet expectations--and its role in evaluation and
decision-making (e.g., Bell, 1985; Gul, 1991; Loomes \& Sugden, 1986).
These researchers have generally argued that expectations modify
people's reactions to events, so that they are jointly determined by the
context-free utilities of options or events (e.g., economic utilities),
and the contextually-dependent disappointment that individuals
experience as a function of their expectations. Similarly, Mellers and
colleagues (1997) interpreted the divergence they observed between the
utility of an outcome (i.e., monetary gains or losses) and the felt
experience of that outcome as a result of post-decision affect.

Each of these theories has been rigorously defined and supported within
the small-world confines they were meant to capture. However, we seek to
craft a more general explanation of how and why expectations influence
evaluations and the range of contexts in which they should be expected
to do so. Several accounts of utility evaluation, including early
theories like Prospect Theory (Kahneman \& Tversky, 1979; Tversky \&
Kahneman, 1992), have emphasized the role of relative comparisons in
evaluations of utility. In a similar spirit, we argue that expectations
can set the reference points against which people compare future
outcomes. On this view, evaluating the utility of some event consists in
comparing the state of the world now that the event has occurred to the
state of the world just prior to the event. Further, we assume that our
knowledge of the world is uncertain, so that it is really just our
(probabilistic) expectations about what has happened, what is happening,
and what will happen in the future. Thus, the evaluation of an event
might be driven by what is learned about the state of the world as a
result of that event's occurrence.

The account outlined above provides an information-theoretic route to
understanding how expectations directly influence evaluation. A
fundamental insight of Information Theory is that the information
carried by an event is a function of its prior probability. This means
that low probability events carry more information than high probability
events (Shannon, 1948). Along similar lines, violations of expectations
have long been recognized as fundamental to associative and animal
learning models; a larger delta means people learn more (e.g., Rescorla
\& Wagner, 1972). Science progresses most when new findings impugn
widely-accepted theories or when a theory's extreme predictions are
validated. Likewise, people learn more about the state of the world when
their expectations are violated by shocking world events as compared to
when they are affirmed by less surprising events. For example, if a
bombing occurs in Paris -- an unexpected location -- rather than Lebanon
--a more expected location--we learn that the world is more dangerous
than we had previously believed.

A more formal treatment of these issues reveals the generality of these
conclusions. Formally, before some binary event \(x_i\) (which either
occurs or does not occur), an evaluator has a mental model of the
current state of affairs, S. This mental model includes expectations
about potential future events and uncertainty over current and past
events. Thus S represents a joint probability distribution over possible
states of affairs \(p(x_1, x_2, ..., x_n)\). The evaluator's subjective
utility function can assign an expected utility to S--representing how
positive or negatively the evaluator views the current state of affairs.
The occurrence of \(x_i\) brings about some new state of affairs, S', to
which the evaluator can again apply her utility function. The evaluation
of event \(x_i\) then consists in the comparison between these two
states of affairs, those before the event (reflecting the agents'
expectations), and the new state of affairs brought about by the event:

\[ V(x_i) = EU(S') -EU(S) \]

We let S be represented by a random vector X, whose members are any
number of discrete events \(x_1, x_2, ..., x_n\), where \(x_i\)
represents the event of interest so that \(x_i\) occurs in S with some
probability and occurs in S' with probability 1. That is, S represents
\(X\) and S' represents \(X|x_i=1\). We write this as:

\[
\begin{aligned}
V(x_i) &= EU(X|x_i) - EU(X) \\
&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x)
\end{aligned}\]

Applying the chain rule of probability and some algebra (see appendix)
we eventually obtain:

\[V(x_i) = \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i)\]

Thus, the value assigned to event \(x_i\) is proportional to the prior
probability of \(x_i\). Further, this holds true regardless of the
nature of the event in question and of the form of the utility function
over events.

In the context of our reaction to surprise parties, it might seem
desirable that expectations would influence our experience of events.
Much joy is tied to the psychological marriage between surprise and
evaluation. However, this basic cognitive process may have negative
moral repercussions. When harms are expected, these dynamics may lead
people to draw less extreme evaluations, feel less concern for victims,
and be less driven to help. To test this hypothesis, we examined whether
people's evaluations of morally harmful events are affected by their
expectations about those events. We asked people to compare pairs of
simple events where a victim suffered an identical harm, but where the
events differed in their prior probability. For each pair of events,
participants were asked to judge which event was more upsetting. Two
studies provided evidence for the predicted effect of expectations on
moral evaluations: people tended to view unexpected negative events as
more upsetting, even when the harm to victims was identical.

\section{General Methods}\label{general-methods}

Here we present two preregistered studies examining the role of
expectations in moral evaluations. Study 1 used a forced-choice task to
test the hypothesis that people would be more upset about moral outcomes
that were unexpected than expected. In Study 2, we tested of our
hypothesis using a more conservative judgment task to increase the
generalizability of our results.

\subsection{Materials}\label{materials}

In both studies, participants were presented with a series of trials
where they read brief (one sentence) descriptions of two different
events and were asked to indicate which of the two events seemed more
upsetting. In ``experimental'' trials, the two events were highly
similar, but differed in their prior probabilities: one event was more
expected and one more unexpected. (The perceived likelihood of a given
event was confirmed in prior norming studies). These prior expectations
were manipulated by changing the context in which the events occurred.
For example, participants considered the following stimulus:

@@

In each event, the harm to the victim is the same (here, death) but one
event is more expected than the other, given the different likelihoods
of earthquakes occurring in California versus Oklahoma.

Each study contained between 16 experimental event-pairs that spanned a
variety of different events and contexts. All experimental materials for
these studies are available as supplemental online materials at {[}OSF
LINK{]}.

Both studies included ``equivalent'' filler trials. In these trials, the
two events differed in trivial contextual details that we did not expect
would affect participants' judgments. For example:

\begin{itemize}
\item
  ``A man in Connecticut starts a house fire.'' {[}Equally expected{]}
\item
  ``A man in New Hampshire starts a house fire.'' {[}Equally expected{]}
\end{itemize}

These filler trials were meant to prevent participants from becoming
explicitly aware of the structure of the experimental trials. Finally,
both studies included ``non-equivalent'' filler trials, the two events
differed substantially in the degree of harm suffered by a victim, so
that one event was expected to be seen as considerably more upsetting
than the other. For example:

\begin{itemize}
\item
  ``An 11-year-old child sets a doll on fire'' {[}Less severe{]}
\item
  ``A 12-year-old child sets a cat on fire'' {[}More severe{]}
\end{itemize}

These trials were included to allow participants a chance to use the
extremes of the response scale and to reduce any task demands that might
drive them to make artificially fine-grained distinctions between the
severity of events.

\subsection{Exclusions}\label{exclusions}

In both studies, we excluded participants who failed to incorrectly
answer attention-check questions. These questions asked participants to
enter a particular response to ensure that they were paying attention
and reading the items as they proceeded through the study. A final
question asked participants if they had paid attention and taken the
study seriously, encouraging them to be honest in their replies.

\subsection{Data Analysis}\label{data-analysis}

We analyzed our data by performing Bayesian estimation using the
probabilistic programming language Stan (Carpenter et al., 2017). We
tested our predictions by computing Bayes Factors (i.e.~BF01) on the
intercept term of our regression model using the hypothesis function in
the R package brms. Bayes Factors express the ratio of the probability
of data under the null hypothesis to the probability of the data under
an alternative hypothesis. Therefore, larger Bayes Factors indicate that
the data are more likely under the null hypothesis (e.g., that the
intercept is not different from zero) than the alternative hypothesis
(e.g., that the intercept is different from zero), and vice versa. Bayes
Factors can be influenced by prior choices (Gelman, Simpson, \&
Betancourt, 2017) so we also performed prior robustness checks to
confirm that our conclusions were unchanged under different prior
assumptions.

\section{Appendix}\label{appendix}

\[ \begin{aligned}
V(x_i) &= EU(X|x_i) - EU(X) \\
&=  \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x) \\
&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x|x_i)p(x_i) \\
&= \sum_{x \in X} U(x)p(x|x_i) - p(x_i)\sum_{x \in X} U(x)p(x|x_i) \\
&= \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i) \\
\end{aligned}\]

The first step utilizes the Law of the Unconscious Statistician, the
next comes from the chain rule of probability, and the final two steps
are simple algebra.

\section{References}\label{references}

\setlength{\parindent}{-0.25in} \setlength{\leftskip}{0.125in} \noindent

\hypertarget{refs}{}
\hypertarget{ref-Bell1985}{}
Bell, D. (1985). Disappointment In Decision Making Under Uncertainty.
\emph{Operations Research}, \emph{33}(1), 1--27.
\url{http://doi.org/10.1287/opre.33.1.1}

\hypertarget{ref-Brewer1999}{}
Brewer, M. B. (1999). The psychology of prejudice: Ingroup love or
outgroup hate? \emph{Journal of Social Issues}, \emph{55}(3), 429--444.
\url{http://doi.org/10.1111/0022-4537.00126}

>>>>

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}

%% Some pieces required from the pandoc template
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\templatetype{pnasresearcharticle}  % Choose template

\title{Expectations bias moral evaluations}

\author[a,1]{Derek Powell}
\author[b]{Zachary Horne}

  \affil[a]{Stanford University, Department of Psychology, 450 Serra Mall, Stanford,
CA, 94305}
  \affil[b]{Arizona State University, School of Social and Behavioral Sciences,
13591 N 47th Ave, Phoenix, AZ 85051}

% Please give the surname of the lead author for the running footer
\leadauthor{Powell}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{People living in geo-politically unstable regions or in the developing
world are often those who are most affected by terrorism, famine, and
natural disasters, and thus are the very people in greatest need of
assistance and concern from the world at-large. However, for these very
reasons, it is often unsurprising when harm befalls them. Two
experiments confirm this intuition: we found that people's expectations
bias their reactions to morally-significant events--people are more
upset by surprising harms than more expected harms. This bias threatens
to reduce observer's concern for victims most in need, making people
less likely to donate time or money to aid victims, or to take political
action to prevent further harms.}

\authorcontributions{Please provide details of author contributions here.}

\authordeclaration{Please declare any conflict of interest here.}

\correspondingauthor{\textsuperscript{} }

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
 \keywords{  evaluation |  moral judgment |  information theory   }

\begin{abstract}
When we learn of tragic events--a violent killing or a terrorist
attack--we are mournful and often driven to action. But our responses to
these events can differ substantially depending on the context of the
event. When tragedy takes us by surprise, we are shocked and outraged,
but when it strikes in unstable regions or rough neighborhoods, we are
often resigned to feel that ``these things happen.'' Here we explain the
role that expectations seem to play in moral evaluation by modeling
evaluations as the comparison of the state of the world before and after
an event. According to this model, expectations influence evaluations
through information-theoretic principles: less expected events do more
to inform us about the state of the world than do more expected events.
Consistent with this theory, in two preregistered studies we found that
people's judgments of morally significant events were affected by the
prior probability of that event, so that people were more upset about an
event that was unexpected (e.g., a robbery at a clothing store) than an
event that was more more probable (e.g., a robbery at a convenience
store). This bias threatens to impose a pernicious cycle: People living
in poverty, geo-politically unstable regions, and other dangerous
circumstances have the greatest need for assistance and concern.
However, for these very reasons, it is often unsurprising when they are
harmed. Consequently, the influence of expectations on moral evaluations
threatens to reduce observer's concern for those people who are most in
need.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.

\acknow{We would like to acknowledge the help of Larisa Hussak, Keith Holyoak,
Alan Fiske, Hongjing Lu, John Hummel, Andrei Cimpian, and Ellen Markman
for their comments and support.}

When we learn of tragic events--a friend's loss of a loved-one, a
violent killing, or a terrorist attack--we are mournful, concerned,
angry, and driven to action. But our responses to these events differ
depending on the context (1). In particular, our expectations play an
important role in our reactions to events. This seems true of both the
mundane--when a friend's loved-one passes, we often ask, ``was it
sudden?''--the dramatic--a murder in a sleepy town prompts a very
different reaction than a similar crime in a major city--and even the
catastrophic: acts of terrorism seem to evoke more outrage, horror, and
concern when they occurs in peaceful rather than unstable regions. For
instance, the 2015 Paris attacks were felt around the world. Yet, most
of those mourning had little to say 15 hours earlier, when another
tragic attack killed at least 43 people in Beirut (2). Likewise, stories
of the Virginia Tech massacre, which killed 32 people, wildly
overshadowed coverage of attacks killing nearly 200 people in Iraq--one
of the bloodiest days since the 2003 invasion (3). In each of these
situations, the surprise elicited by an event appears to affect people's
reactions to the event. In some contexts, harm is tragic and in others,
``these things happen.''

Although this pattern seems so intuitive as to be obvious, no reasonable
code of ethics would justify this pattern of judgment, nor do any moral
psychological frameworks explicate the role of expectations in moral
evaluations. Outside of the domain of morality, however, everyday
experiences and controlled experiments show that people's evaluations of
events unambiguously depend on their expectations about those events.
There is a special thrill to having one's expectations exceeded and
disappointment when events fail to meet expectations: we root for the
underdog, hold surprise parties, and foreshadow bad news to ease its
delivery (4). Indeed, in a controlled study, Mellers and colleagues (5)
found that expectations influenced affective reactions during a gambling
task: Given a gamble with a 10\% chance to win \$30 and 90\% chance to
win \$0, participants felt little disappointment with the \$0 outcome,
but were considerably more excited when they won \$30. Conversely, given
a gamble with a 90\% chance of winning \$60 and 10\% chance of winning
\$0, participants were disappointed with the \$0 outcome and showed more
muted enjoyment of the \$60 outcome. In fact, in gambles similar to
these, people were happier with the smaller unexpected gain than with
the larger but more expected gain (6).

Applied to moral issues, the influences of expectations on evaluation
would be a dangerous bias. Tragedies are tragic, whether expected or
not, and failure to recognize that could lead to tacit endorsement of
the pain and suffering of others wherever pain and suffering are the
status quo. Do expectations directly bias reactions to morally
consequential events, as intuition and anecdote seem to suggest? To
demonstrate the plausibility of this suggestion, we develop a
theoretical account of how expectations influence evaluations, not as a
strange quirk or uniquely human bias, but instead as a fundamental
component of how people evaluate events.

Most events of consequence consist of changes--a transition from one
state to another. For this reason, we construe event evaluation as a
comparison between the states of the world before and after the event.
Comparison plays a crucial role in many theories of decision-making, and
especially in theories aimed at explaining the role of expectations in
evaluation (4, 5, 7, 8). For instance, in Prospect Theory (9, 10), the
utility of a prospect is evaluated by comparison to a current reference
point. Our proposal is similar: we argue that expectations can set the
reference points against which people compare future outcomes, so that
assigning value to an event consists in comparing the state of the world
following the event to the state of the world just prior to the event.
As knowledge of the world is uncertain, our understanding of these
states is probabilistic--that is, our model for the state of the world
is just our expectations about what has happened, what is happening, and
what will happen in the future. One direct implication of this proposal
is that the evaluation of an event is linked to what is learned about
the world as a result of that event's occurrence. That is, we evaluate
events positively or negatively to the extent that they inform our
understanding of the world in positive or negative ways.

This account thereby provides an information-theoretic route to
understanding how expectations directly influence evaluation. A
fundamental insight of Information Theory is that the information
carried by an event is a function of its prior probability; low
probability events carry more information than high probability events
(11). We propose that people learn more about the state of the world
when their expectations are violated by shocking world events as
compared to when they are affirmed by less surprising events. For
example, if a bombing occurs in Paris--a less expected location--rather
than Lebanon--a more expected location--we learn that the world is more
dangerous than we had previously believed. This suggestion is novel in
the moral domain, but these points are widely accepted elsewhere. For
instance, expectations have long been recognized as fundamental to
models of human and animal learning (12). Similarly, science progresses
most when new findings impugn widely-accepted theories or when a
theory's extreme predictions are validated.

A more formal treatment of this theory establishes the potential for its
generality across domains. Let the state of the world prior to some
event be represented by a random variable, \(X\). We'll assume \(X\) is
discrete, and that every realization \(x\) represents some potential
specific state of the world in all relevant aspects under consideration.
An agent has some utility function \(u\), that applies over the
different realizations of \(X\). Thus the expected utility of the
present state of the world is:

\begin{equation}
E_X[u(X)] = \sum_x u(x)p(x)
\end{equation}

When we learn an event has occurred, we can represent this as an
observation of a second, binary random variable, \(S\). We can think of
\(S\) as a ``signalling event'' that informs us about about \(X\) (13).
That is, we assume there is a dependency between \(S\) and \(X\), so
that observing \(S\) informs us about the state of the world, \(X\). For
instance, reading a news report of a terror attack in Brussels informs
us about the state of the world, telling us about what realizations of
\(X\) are no longer possible (e.g., any wherein it was a peaceful day in
Brussels) and which are more or less likely. Conditioned on \(S\), the
expected utility of the world is:

\begin{equation}
E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
\end{equation}

Given our construal of event evaluation as the comparison between states
of the world following and prior to the event, we define the value
assigned to the signalling event, \(V(S)\), as:

\begin{equation}
V(S) = E_{X|S}[u(X)] - E_X[u(X)]
\end{equation}

From this it can be shown that the value assigned to event \(S\) is
proportional to its prior probability (see Theoretical Formalization).
Specifically, it is the difference in the expected utility under
different values of \(S\), weighted by the inverse of the prior
probability of \(S\)..\footnote{It should be recognized that the result
  we obtain here bears a similarity to the ``disappointment function''
  included in Mellers and colleagues' (5) Decision Affect Theory.}

\begin{equation}
V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
\end{equation}

This conclusion does not depend on the nature of the event in question
nor on the utility function over events, but instead holds generally
wherever we conceive of evaluation as a holistic comparison between
uncertain states of the world. Consequently, this account predicts
expectations will equally affect reactions to gambles with clearly
specified values and probabilities (5) and real-world events with
ill-defined values and where reasoners' expectations are defined by
their existing beliefs and world knowledge.

Given the potentially fundamental role expectations play in event
evaluation, and the potentially serious moral biases that may result, we
sought to test empirically whether people's evaluations of morally
harmful events are affected by their expectations about those events. We
conducted two preregistered studies examining the role of expectations
in moral evaluations. In both studies, participants were presented with
a series of trials where they read brief descriptions of two different
events and were asked to indicate which of the two events seemed more
upsetting. In these trials, the two events were highly similar, but were
manipulated so that they differed in their perceived prior
probabilities: one event was more expected and one more unexpected.

The design of these studies was chosen with special consideration to two
theoretical points. First, the information-theoretic account we have
described makes a solely directional prediction--all else being equal,
more surprising events will elicit stronger reactions than less
surprising events. Asking participants to compare two similar events
allowed us to test this directional prediction. Second, it is the
informativeness of events, determined by one's prior expectations of the
event, that we predict will shape evaluations. Thus, reasoners should
not have to be induced to hold expectations experimentally (5), but
instead their own prior beliefs will be sufficient to shape their
expectations and affect their evaluations. For this reason, we
manipulated the prior probability of events by describing two otherwise
identical events occurring in different contexts. This allowed us to
test the role of expectations based on world knowledge, as well as
providing a more naturalistic set of items.

\begin{figure}
\centering
\includegraphics{fig1.eps}
\caption{Probability with which participants chose the less expected
event as more upsetting across trials in Study 1. As predicted,
participants exhibited a robust bias toward choosing the less expected
event as more upsetting across all items. Error bars represent 95\%
credible intervals calculated from the hierarchical logistic regression
model. See Supporting Information for full description of items. {}}
\end{figure}

Consistent with our predictions, manipulating expectations consistently
shaped evaluations across a range of morally-significant events. We
found that people viewed unexpected harmful events as more upsetting
than expected harmful events, even when the harm a victim suffered was
the same. For instance, people judged that a robbery at a clothing store
(unexpected) was more upsetting than a robbery at a convenience store
(expected). Furthermore, these differences were large, systematic, and
robust. Although there was variability in the situations that elicited
the largest differences, our statistical approach treated trial as a
random effect (see Methods and Results), allowing us to generalize
conclusions to items we did not test (14).

\begin{figure}
\centering
\includegraphics{fig2.eps}
\caption{Participants' responses by item in Study 2. Upper: heatmap plot
displaying probability of choosing each response on the 1 to 5 scale
across items. Lower: marginal expected response across items, with 95\%
credible interval calculated from the ordinal regression model. Note the
scale has been constrained to show relevant region of results. In both
plots, response of 3 indicates indifference, a response below 3
indicates a bias to choose the more expected event as more upsetting,
and a response above 3 indicates a bias to choose the less expected
event as more upsetting. Similar to Study 1, when participants judged
that one of the events was more upsetting, they showed a robust bias
toward choosing the less expected event as more upsetting. Trials are
sorted by response to aid readability. {}}
\end{figure}

We have argued that expectations shape our reactions to events because
they determine how informative those events are in shaping our
understanding of the world. Thus, we argue that well-grounded
information theoretic principles can account for people's divergent
moral reactions to likely and unlikely events.

There's a sense in which it is clearly adaptive that the violation of
our expectations would direct our attention to the events and situations
about which our understanding is most in need of revision. For instance,
an act of terrorism in a peaceful nation might be less expected because
of our beliefs about the effectiveness of security forces, the stability
of national relationships, and positive relationships among people
groups in that nation. Therefore, a surprising act of terrorism casts
doubt on our deeper understanding of the geopolitical situation, and
thereby leads us to suspect similar acts are more likely to occur in the
future.

Yet, however well-grounded the underlying principles, there's a clear
moral danger whenever expectations influence our moral concerns. For
instance, ``victim blaming'' describes the tendency to shift blame away
from aggressors and onto victims of crimes like sexual assault (15) and
police violence. Work by Janoff-Bulman and colleagues (16) has shown
that merely increasing the perceived probability of a victim suffering a
sexual assault induced participants to blame the victim for her assault.
People's expectations may explain the flip side of victim blaming,
whereby those who victim blame also tend to judge transgressors less
harshly: When people perceive a victims' actions as increasing the
probability of their eventual assault, they may perceive an aggressor's
actions as producing a smaller change in utility, thus warranting weaker
condemnation.

The influence of expectations on moral evaluations poses perhaps the
greatest threat when considering our reactions to humanitarian
challenges brought on by poverty or regional instability. In these
cases, the influence of expectations on moral evaluations threatens to
impose vicious and morally pernicious cycles. For instance, people
living in geo-politically unstable regions or in the developing world
are often those who are most affected by terrorism, famine, and natural
disasters, and are the very people in greatest need of assistance and
concern from the world at-large. However, for these very reasons, it is
often unsurprising when harm befalls people living in these
circumstances. Thus the influence of expectations on moral evaluations
threatens to reduce observer's concern for these victims most in need,
making people less likely to donate time or money to aid victims, or to
take political action to prevent further harms. The present research
identifies this potentially pernicious moral bias and grounds it in
information theoretic principles. Hopefully, by further understanding
these cognitive dynamics, psychologists and policymakers can begin to
find ways to reduce or counteract people's tendency to ignore the plight
of those in most need of attention.

\section*{Materials and Methods}\label{materials-and-methods}
\addcontentsline{toc}{section}{Materials and Methods}

We conducted two preregistered studies examining the role of
expectations in moral evaluations using two different judgment tasks
(\url{https://osf.io/5qnf7/}). Each study consisted of multiple trials
where participants were presented brief descriptions of two events.
Their task was to decide which of the two events was more upsetting. In
Study 1, they made a two-alternative forced-choice, and in Study 2 they
made their judgments using a five-point rating scale. Both studies
consisted of 16 experimental trials, five ``equivalent'' filler trials,
and five ``non-equivalent'' filler trials.

In experimental trials, the two events described identical harms, but
one event was more expected and one more unexpected, as determined in a
prior norming study (see SI). For example, participants considered the
following stimulus:

@@

In each of these events, the harm to the victim is the same (death) but
one event is more probable than the other, given the different
likelihoods of earthquakes occurring in California versus Oklahoma. In
addition to being more naturalistic, manipulating expectations through
(otherwise irrelevant) contextual details helps ensure that the harm
suffered by the victim is understood to be identical. In the context of
gambles, it is possible to manipulate probabilities specifically and
directly (5). However, in the context of more realistic moral harms, any
attempt to directly manipulate probabilities risks also changing
participants' perceptions of the degree of harm. For instance, suppose
we describe two cancer patients, one with a 5\% chance of survival and
one with a 50\% chance of survival. Even if they both ultimately meet
the same fate, participants might fairly assume the patient with the
lower chance of survival had a more serious illness, was more
debilitated, suffered more severely, and so forth. In contrast, changing
the context of the two events explains away these likelihood
differences, better equating the events in other respects.

Both studies also included ``equivalent'' filler trials. In these
trials, the two events differed in trivial contextual details that we
did not expect would affect participants' judgments. These filler trials
were meant to prevent participants from becoming explicitly aware of the
structure of the experimental trials. Finally, both studies included
``non-equivalent'' filler trials, the two events differed substantially
in the degree of harm suffered by a victim, so that one event was
expected to be seen as considerably more upsetting than the other. These
trials were included to allow participants a chance to use the extremes
of the response scale and to reduce any task demands that might drive
them to make artificially fine-grained distinctions when generating
their responses. All experimental materials for these studies are
available as Supporting Information at \url{https://osf.io/5qnf7/}.

\subsection*{Exclusions}\label{exclusions}
\addcontentsline{toc}{subsection}{Exclusions}

In both studies, we excluded participants who failed to correctly answer
attention-check questions or who indicated at the end of the study they
had not taken their participation seriously.

\subsection*{Data Analysis}\label{data-analysis}
\addcontentsline{toc}{subsection}{Data Analysis}

We analyzed our data by performing Bayesian estimation using the
probabilistic programming language Stan (17). We tested our
preregistered predictions by computing Bayes Factors (i.e.~BF01) using
the hypothesis function in the R package brms. Bayes Factors express the
ratio of the probability of data under the null hypothesis to the
probability of the data under an alternative hypothesis. Larger Bayes
Factors indicate that the data are more likely under the null hypothesis
(i.e., that the intercept is not different from zero) than the
alternative hypothesis (i.e., that the intercept is different from
zero), and vice versa. As Bayes Factors can be influenced by prior
choices (Gelman, Simpson, \& Betancourt, 2017), we performed prior
robustness checks to confirm that our conclusions would not depend
substantially on the specification of the priors.

To ensure the generality of our findings, we tested participants'
reactions to a variety of items and treated both participants and items
as random effects throughout our analyses. The use of random effects
over items allows us to consider these items as a sample from a larger
population of possible items, and to generalize our findings to that
wider population, much as we generalize from our sample of participants
to the larger population (14).

\subsection*{Study 1}\label{study1}
\addcontentsline{toc}{subsection}{Study 1}

\subsubsection*{Participants}\label{s1-participants}
\addcontentsline{toc}{subsubsection}{Participants}

A total of 255 participants were recruited from Amazon's Mechanical Turk
work distribution website (mTurk). Of these, 230 passed attention checks
and were included in the final analyses (105 male, 125 female, median
age = 33 years old). All participants were paid \$0.60 for their
participation.

\subsubsection*{Materials and procedure}\label{s1-materials-procedure}
\addcontentsline{toc}{subsubsection}{Materials and procedure}

Participants judged 16 experimental event-pairs and 10 equivalent filler
event-pairs. On each trial (an example is shown above), participants
were presented with the event-pair stimulus and had to judge which
outcome was more upsetting in a two-alternative forced choice task. The
two events were labeled ``Outcome 1'' and ``Outcome 2'' and the order,
and whether Outcome 1 or 2 was the unexpected event, was randomized and
counter-balanced.

\subsubsection*{Results and discussion}\label{s1-results}
\addcontentsline{toc}{subsubsection}{Results and discussion}
